`r params$doc_title`

### 2022-11-12 scripts/MSc.Rmd
updated figures, figure01 quadrant function

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(ggplot2)
library(scales)
library(ggthemes)
#library(biomaRt)  # different script
library(kableExtra)
library(MendelianRandomization)
library(mr.raps)
library(MRMix)
library(MRPRESSO)
library(penalized)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```

```{r}
my_data <- as_tibble(read.delim("../data-boekstegers_felix/GST_BMI_association_results.txt"))
# GBC: new data 2022-11-04
GBC_dat <- as_tibble(read.delim("../data-boekstegers_felix/C23C24_BMI_association_results.txt"))

GBC_dat <- GBC_dat %>% 
  rename(eaf_gbc=eaf_out,
         beta_gbc=beta_out,
         se_gbc=se_out,
         p_value_gbc=p_value_out) %>% 
  select(-c(chr, pos_hg19, pos_hg38, a1, a2, exposure, outcome))
my_data <- my_data %>% left_join(GBC_dat, by =c("SNP", "dummyID"))
```

# Get gene names from rs-ids

```{r}
# see script biomart_search.R
rsIds_GeneNames <- as_tibble(readRDS("../data/biomartBMIsearch.rds"))

filter_FTO <- rsIds_GeneNames %>% 
  filter(str_detect(associated_gene, pattern = "FTO")) %>% 
  distinct(refsnp_id, .keep_all = TRUE)
```

@@

## Calculate approximate SEs for decreasing sample sizes and corresponding p-vals

Calculating p-values from a normal distribution. The se, pval and study size depend on each other, within each study. The selection of IVs should rely on the exposure study - BMI in our case. Summary statistics on genetic associations with BMI will be based on results of the GIANT consortium + GERA (n=334487 (100,418 + 234,069) individuals and 289 associated variants with p < 10-8) **(insert citation)**. We vary the sample size and calculate the number of IVs and the explained variance. *(Justo Mail 2022-09-27)*
\
Steps:
1.       Calculate more accurates SEs using the betas and Pvals provided to you (this will be the original/initial SE for n=334.487). 
2.       Calculate the approximate SEs for decreasing samples sizes
3.       Calculate the probability values considering the estimated betas and the corresponding SEs
4.       Exclude the IV if Pval > genome-wide significance threshold  (5 × 10−8) **NEW 5x10^(-6))**
5.       Calculate cumulative explained variances
6.       Represent figure similar to Fig1 in Chatterjee’s paper

For the selected IVs, please also calculate the explained variance for the confounder and outcome.

```{r z-statistic}
### justo approach email 30-10-2022  ---------------------------------------

## LOOP simulation ----------------------------------------------
n.orig <- 334487
n <- c(1000, 2000, 5000, seq(10000, 330000, by=10000), 334487)

# create vectors to store pvals_n in
colnames_pval <- c()
for (i in 1:length(n)) {
  colnames_pval[i] <- paste0("p_value_exp_sample_", format(n[i], scientific = FALSE))
}
## initialize
se_update <- c()
pvals_df <- data.frame(matrix(ncol = length(colnames_pval), nrow =  nrow(my_data)))
colnames(pvals_df) <- colnames_pval
for (i in 1:nrow(my_data)) {
  # 1) Calculate more accurates SEs using the provided betas and Pvals 
  se_update[i] <- my_data[["beta_exp"]][i]/-qnorm(p=(my_data[["p_value_exp"]][i]/2),mean=0)

@@

  # 3) Calculate the probability values considering the estimated betas and the corresponding SEs
  for (j in 1:length(n)) {
    pvals_n <- 2*(1-pnorm(abs(my_data[["beta_exp"]][i])/abs(se_n[j]), mean=0, 
                          lower.tail = TRUE))
    pvals_df[i, j] <- pvals_n
  }
}

@@

my_data <- as_tibble(cbind(my_data, pvals_df))

@@

@@

@@

## calculate explained variance
my_data <- my_data %>% 
  mutate(explained_variance_BMI = explained_variance_numeric(eaf = eaf_exp, beta = beta_exp), 
         explained_variance_GSD = explained_variance_binary(PA = eaf_out, 
                                                            RR1 = exp(beta_out), 
                                                            RR2 = exp(beta_out)^2, 
                                                            K = 0.1  # assumed prevalence in population
                                                            )$Vg,
         explained_variance_GBC = explained_variance_binary(PA = eaf_gbc, 
                                                            RR1 = exp(beta_gbc), 
                                                            RR2 = exp(beta_gbc)^2, 
                                                            K = 1.2/100000  # assumed prevalence in population
                                                            )$Vg
         )

## paper backing prevalence GSD
# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3343155/#:~:text=Gallstones%20are%20common.,develop%20stones%20at%20some%20time.&text=The%20majority%20will%20not%20develop,cholecystitis%2C%20cholangitis%2C%20or%20pancreatitis
## paper backing prevalence GSD (ASR/100.000 = 1.2 worldwide)
# https://www.wcrf.org/cancer-trends/gallbladder-cancer-statistics/

#explained_variance_binary(PA=0.3,RR1=1.3,RR2=1.3^2,K=0.01)
```

@@

```{r}
# labeling (only FTO label)
my_data$gene_label <- NA
my_data$gene_label[my_data$SNP %in% filter_FTO$refsnp_id] <- "FTO"

# color labeling
my_data$fto_col_annot <- "No FTO variant"
my_data$fto_col_annot[my_data$SNP %in% filter_FTO$refsnp_id] <- "FTO variant"

ggplot(data = my_data, 
       mapping = aes(x=explained_variance_BMI*100, y= -log10(p_value_exp),  
                    col=fto_col_annot, label=gene_label)
       )+
  geom_point(alpha=0.6)+
  ggrepel::geom_text_repel(show.legend = FALSE)+
  xlab("Explained variance in BMI in %")+
  ylab("-log10 P-val")+
  labs(title = "Explained variance exposure",
       col='SNP'
       ) +
  theme_few()+
  scale_colour_few() -> fig00
fig00

ggsave(filename = "../output/plots/figure00.png", plot = fig00, 
       width = 10, height = 6, 
       units = "in"  # default
       )
```

@@

```{r get-data-for-fig1}
# 4.       Exclude the IV if Pval > genome-wide significance threshold  (5 × 10−8) **NEW 5x10^(-6))**
# 5.       Calculate cumulative explained variances
# initialize df
df_cum_expl_var <- tibble(n=n, 
                          cum_expl_varX = rep(NA, length(n)), 
                          cum_expl_varU = rep(NA, length(n)), 
                          cum_expl_varY = rep(NA, length(n)),
                          count_IVs = rep(NA, length(n))
)

for (i in 1:length(colnames_pval) ) {
  threshold <- 5*10^(-6)
  select_col <- colnames_pval[i]
  # calculate cum explained variance and count of IVs
  summarydat <- my_data %>% 
    filter(!!sym(select_col) < threshold) %>% 
    select(explained_variance_BMI, explained_variance_GSD, explained_variance_GBC) %>% 
    summarise(expl_variances =across(.cols = everything(), function(x) sum(x)),
              count_IVs = n())
  # fill up df
  df_cum_expl_var[i, 2] <- summarydat$expl_variances[1]
  df_cum_expl_var[i, 3] <- summarydat$expl_variances[2]
  df_cum_expl_var[i, 4] <- summarydat$expl_variances[3]
  df_cum_expl_var[i, 5] <- summarydat$count_IVs
}
```

```{r plot-fig1}
# 6.       Represent figure similar to Fig1 in Chatterjee’s paper

# Dual Y axis: we have 284 IVs, scale them to 5% (y-axis [0;5%]): 3/284 
## https://r-graph-gallery.com/line-chart-dual-Y-axis-ggplot2.html
scale_factor <- 4.5/284
df_cum_expl_var <- df_cum_expl_var %>% 
  mutate(
    cum_expl_varX = cum_expl_varX *100, # in percent 
    cum_expl_varU = cum_expl_varU *100,
    cum_expl_varY = cum_expl_varY *100,
    count_IVs_scaled = count_IVs*scale_factor)

plot_data <- df_cum_expl_var %>% 
  pivot_longer(cols = c(cum_expl_varX, cum_expl_varU, count_IVs_scaled)) %>% 
  mutate(name = factor(name, labels = c("Number of IVs", "% var(U)", "% var(X)")))

ggplot(plot_data, aes(x=log10(n), y=value, col=name))+
  geom_point(alpha=0.6)+
  geom_line(alpha=0.6)+
  xlab("log10(N)")+
  #ylab("% variance explained by IVs")+
  labs(title = "Figure 1",
       col='', 
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  scale_y_continuous(
    # Features of the first axis
    name = "% variance explained by IVs",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./scale_factor, name="Number of IVs")
  ) + 
  theme_few()+
  scale_colour_few() ->fig01
fig01

ggsave(filename = "../output/plots/figure01.png", plot = fig01, 
       width = 10, height = 6, 
       units = "in"  # default
       )
```

```{r updated-fig01}
plot_data1 <- df_cum_expl_var %>% 
  pivot_longer(cols = c(cum_expl_varX, cum_expl_varU, cum_expl_varY)) %>% 
  mutate(name = factor(name, labels = c("% var(U)", "% var(X)", "% var(Y)")))

N_annotation <- c(10000, 100000, 200000, 300000)
## plot
ggplot(plot_data1, aes(x=n, y=value, col=name))+
  geom_point(alpha=0.6)+
  geom_line(alpha=0.6)+
  geom_line(aes(x=n, y=count_IVs_scaled, col="Number of IVs"), linetype="dotdash")+
  scale_x_continuous(name = "Sample Size",
                     labels = comma, 
                     breaks = N_annotation, #c( 10000, n[7:length(n)]),
                     limits = c(10000, 350000))+
  #ylab("% variance explained by IVs")+
  labs(title = "Figure 1",
       col='', 
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  scale_y_continuous(
    # Features of the first axis
    name = "% variance explained by IVs",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./scale_factor, name="Number of IVs")
  ) + 
  theme_few()+
  #scale_colour_few() +
  #scale_color_manual(name = "", values = c("% var(U)" = "red", 
  #                                         "% var(X)" = "blue", 
  #                                         "% var(Y)" = "darkblue", 
  #                                         "Number of IVs" = "black"
  #                                         ))
  #
  scale_color_manual(name = "", values = c("% var(U)" = ggthemes_data$few$colors$Dark[2,2][[1]], 
                                           "% var(X)" = ggthemes_data$few$colors$Dark[3,2][[1]], 
                                           "% var(Y)" = ggthemes_data$few$colors$Dark[4,2][[1]], 
                                           "Number of IVs" = "black"
                                           )) -> fig01.update
fig01.update
# scale_color_manual(name = "Y series", values = c("Y1" = "darkblue", "Y2" = "red"))
ggsave(filename = "../output/plots/figure01update.png", plot = fig01.update, 
       width = 10, height = 6, 
       units = "in"  # default
       )
```

@@

```{r MAF-Pval}
ggplot(data = my_data, aes(x=eaf_exp, y=-log10(p_value_exp)))+
  geom_point(alpha=0.6)+
  geom_hline(yintercept=-log10(median(my_data$p_value_exp)), col="red", alpha=0.6)+
  geom_vline(xintercept=median(my_data$eaf_exp), col="red", alpha=0.6)+
  xlab("Effect allele frequency for exposure")+
  ylab("-log10(P-value) ")+
  labs(title = "Figure 2",
       subtitle = "Define different genetic architectures",
       #col='', 
       #caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  theme_few()+
  scale_colour_few() ->fig02
fig02

ggsave(filename = "../output/plots/figure02.png", plot = fig02)
```

@@

```{r mafVSbetas}
plot_data_2 <- my_data %>% 
  mutate(maf_exp = ifelse(eaf_exp >0.5, 1-eaf_exp, eaf_exp)) %>% 
  select(maf_exp, beta_exp)

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
figurespath <- paste0("../reports/", format(Sys.time(), "%Y%m%d"), '_figures/')
knitr::opts_chunk$set(fig.path = figurespath)  # reference is script! also creates new dirs

library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(ggplot2)
library(scales)
library(ggthemes)
#library(biomaRt)  # different script
library(kableExtra)
library(MendelianRandomization)
library(TwoSampleMR)
library(mr.raps)
library(MRMix)
library(MRPRESSO)
library(penalized)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
```

# Discussion

-   2022-11-12:
    -   8 SNPs removed after data harmonization, one SNP associated with FTO gene (none left)
    -   in new formula for proportion of variance in phenotype explained by given SNP (PVE) we need the MAF not the EAF
        - calculate: if $eaf>0.5$ -> $maf = 1-eaf$
        - need to change betas as well? +-
    -   We discussed $K=0.05$ in our last meeting for the "prevalence" of GBC after looking at [Cancer Today](https://gco.iarc.fr/today/home) 
        - If done this way we yield a 20% proportion of variance explained in PVE. Switch back to ASR used last week!

# Version Update

The new report version includes:

- First analysis

```{r load-data}
my_data <- as_tibble(read.delim("../data-boekstegers_felix/GST_BMI_association_results.txt"))
# GBC: new data 2022-11-04
GBC_dat <- as_tibble(read.delim("../data-boekstegers_felix/C23C24_BMI_association_results.txt"))
```

# Data Harmonization

We start with harmonizing our data sets for the exposure

```{r data-harmonization, message=FALSE}
##exposure_dat:
#SNP
#beta.exposure
#se.exposure
#effect_allele.exposure
#other_allele.exposure
#eaf.exposure
exposure_dat <- my_data %>% 
  select(dummyID, SNP, beta_exp, se_exp, a1, a2, eaf_exp, exposure) %>% 
  rename(
    id.exposure = dummyID,
    beta.exposure = beta_exp,
    se.exposure = se_exp,
    effect_allele.exposure = a1,
    other_allele.exposure = a2, 
    eaf.exposure = eaf_exp, 
    exposure = exposure
  )
##outcome_dat:
#SNP
#beta.outcome
#se.outcome
#effect_allele.outcome
#other_allele.outcome
#eaf.outcome
#outcome
outcome_dat <- GBC_dat %>% 
  select(dummyID, SNP, beta_out, se_out, a1, a2, eaf_out, outcome) %>% 
  rename(
    id.outcome = dummyID, 
    beta.outcome = beta_out,
    se.outcome = se_out,
    effect_allele.outcome = a1,
    other_allele.outcome = a2, 
    eaf.outcome = eaf_out,
    outcome = outcome
  )

harmonised_dat <- TwoSampleMR::harmonise_data(
  exposure_dat = exposure_dat, 
  outcome_dat = outcome_dat, 
  action = 2  # default
  )

sum_SNPs_remove <- sum(harmonised_dat$remove)
sum_SNPs_palindromic <- sum(harmonised_dat$palindromic)
sum_SNPs_ambiguous <- sum(harmonised_dat$ambiguous)
sum_SNPs_remove <- nrow(harmonised_dat)-sum(harmonised_dat$mr_keep)
```

```{r combine-dat}
my_data_harm <- 
  harmonised_dat %>% 
  rename(
    exposureBMI = exposure, 
    outcomeGBC = outcome,
    # estimates exposure BMI
    eaf.bmi=eaf.exposure,
    beta.bmi=beta.exposure,
    se.bmi=se.exposure,

    # estimates outcome GBC
    eaf.gbc=eaf.outcome,
    beta.gbc=beta.outcome,  
    se.gbc=se.outcome,
  ) %>% 
  left_join(my_data %>% select(SNP, p_value_exp, outcome, p_value_out, eaf_out, beta_out, se_out),by = c("SNP"="SNP")) %>% 
  rename(
    # BMI only need to add p-val
    p.value.bmi = p_value_exp,

    # GSD rename to "confounder"
    confounderGSD = outcome, 
    eaf.gsd = eaf_out,
    beta.gsd = beta_out, 
    se.gsd = se_out, 
    p.value.gsd = p_value_out,
  ) %>% 
  select(
    SNP, 
    # DROP:   "effect_allele.exposure" "other_allele.exposure"  "effect_allele.outcome"  "other_allele.outcome"

    # EXPOSURE
    eaf.bmi,
    beta.bmi,
    se.bmi,
    p.value.bmi,

    # CONFOUNDER
    eaf.gsd,
    beta.gsd,
    se.gsd,
    p.value.gsd,
    # OUTCOME
    eaf.gbc,
    beta.gbc,
    se.gbc,
    # pvalue missing and needs to be added from GBC_dat
    # harmonize decision
    palindromic, ambiguous, mr_keep
  ) %>% 
  left_join(GBC_dat %>% select(SNP, p_value_out), by=c("SNP"="SNP")) %>% 
  rename(
    p.value.gbc = p_value_out
  ) %>% 
  relocate(
    p.value.gbc, .before = palindromic
  ) %>% 
  as_tibble()
```

After harmonizing our data sets we remove `r sum_SNPs_remove` SNPs since they are ambiguous.

```{r rm-ambigous-SNPs}
dropped_data <- my_data_harm %>% 
  filter(mr_keep==FALSE)

my_data_harm <- my_data_harm %>% 
  filter(mr_keep==TRUE)
```

## Get gene names from rs-ids

```{r highlight-fto-df}
# see script biomart_search.R
rsIds_GeneNames <- as_tibble(readRDS("../data/biomartBMIsearch.rds"))

filter_FTO <- rsIds_GeneNames %>% 
  filter(str_detect(associated_gene, pattern = "FTO")) %>% 
  distinct(refsnp_id, .keep_all = TRUE)

t(filter_FTO) %>% kable() %>% 
  kable_styling(font_size=10)
```

After data harmonization we drop `r sum_SNPs_remove` IVs one of which is the SNP associated with the FTO gene.

\clearpage

@@

**Calculate approximate SEs for decreasing sample sizes and corresponding p-vals**

-   se, pval and study size depend on each other, within each study
-   selection of IVs based on exposure study (BMI in our case)
-   Summary statistics on genetic associations with BMI based on results of the GIANT + GERA studies [@Hoffmann_2018] with $N=100,418 + 234,069 =334487$ and $SNPs=289$

**Steps**

1.  Calculate more accurate SEs using the betas and Pvals
2.  Calculate the approximate SEs for decreasing samples sizes
3.  Calculate the probability values considering the estimated betas and the corresponding SEs
4.  Exclude the IV if Pval $\ge$ genome-wide significance threshold (5 × 10−6)
5.  Calculate cumulative explained variances
6.  Represent figure similar to Fig1 in Chatterjee's paper @QiChatterjee_2021

```{r z-statistic}
## LOOP simulation ----------------------------------------------
n.orig <- 334487
n <- c(1000, 2000, 5000, seq(10000, 330000, by=10000), 334487)

# create vectors to store pvals_n in
colnames_pval <- c()
for (i in 1:length(n)) {
  colnames_pval[i] <- paste0("p_value_exp_sample_", format(n[i], scientific = FALSE))
}
## initialize
se_update <- c()
pvals_df <- data.frame(matrix(ncol = length(colnames_pval), nrow =  nrow(my_data_harm)))
colnames(pvals_df) <- colnames_pval
for (i in 1:nrow(my_data_harm)) {
  # 1) Calculate more accurates SEs using the provided betas and Pvals 
  se_update[i] <- my_data_harm[["beta.bmi"]][i]/-qnorm(p=(my_data_harm[["p.value.bmi"]][i]/2),mean=0)

@@

  # 3) Calculate the probability values considering the estimated betas and the corresponding SEs
  for (j in 1:length(n)) {
    pvals_n <- 2*(1-pnorm(abs(my_data_harm[["beta.bmi"]][i])/abs(se_n[j]), mean=0, 
                          lower.tail = TRUE))
    pvals_df[i, j] <- pvals_n
  }
}

@@

my_data_harm <- as_tibble(cbind(my_data_harm, pvals_df))

@@

## Calculate proportion of variance in phenotype explained by a given SNP (PVE)

-   The variance in BMI explained by the IVs is calculated with the following formula: $explained\ variance = 2f(1-f)\beta^2$, where $f$ denotes the allele frequency and $\beta$ is the additive genetic effect.
-   The explained variance in liability to GSD and GBC is calculated as described by @So_2011
    -   Prevalence of disease $K$ must be set:
        -   [Prevalence GSD](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3343155/#:~:text=Gallstones%20are%20common.,develop%20stones%20at%20some%20time.&text=The%20majority%20will%20not%20develop,cholecystitis,%20cholangitis,%20or%20pancreatitis)
            - @Aerts_2003
        -   [Prevalence GBC](https://www.wcrf.org/cancer-trends/gallbladder-cancer-statistics/) 
            - Update prevalence $K=0.05$ from [Cancer Today](https://gco.iarc.fr/today/home)

We added a second formula to calculate the explained variance for numeric traits [@Teslovich_2010].

@@

explained_variance_numeric2 <- function(maf, beta, se_beta, samplesize) {
  ### Teslovich Nature 2010
  # 1) maf = minor allele frequency
  # 2) beta = additive genetic effect
  # 3) se_beta = standard error of beta estimate
  # 4) samplesize = samplesize of study
  maf <- ifelse(maf>0.5, 1-maf, maf)

  explained_variance_numeric <- (2*beta^2*maf*(1-maf))/ 
    ( 2*beta^2*maf*(1-maf) + (se_beta^2)*2*samplesize*maf*(1-maf) )
  return(explained_variance_numeric)
}

@@

## calculate explained variance
my_data_harm <- my_data_harm %>% 
  mutate(explained_variance_BMI = explained_variance_numeric(eaf = eaf.bmi, beta = beta.bmi), 
         explained_variance_BMI2 = explained_variance_numeric2(maf=eaf.bmi, 
                                                           beta=beta.bmi, 
                                                           se_beta=se.bmi, 
                                                           samplesize=n[length(n)]

         ),
         explained_variance_GSD = explained_variance_binary(PA = eaf.gsd, 
                                                            RR1 = exp(beta.gsd), 
                                                            RR2 = exp(beta.gsd)^2, 
                                                            K = 0.1  # assumed prevalence in population
                                                            )$Vg,
         explained_variance_GBC = explained_variance_binary(PA = eaf.gbc, 
                                                            RR1 = exp(beta.gbc), 
                                                            RR2 = exp(beta.gbc)^2, 
                                                            K =1.2/100000  # assumed prevalence in population
                                                            )$Vg
         )
```

@@

```{r PVE_BMI_vs_pval, warning=FALSE}
# labeling (only FTO label)
my_data_harm$gene_label <- NA
my_data_harm$gene_label[my_data_harm$SNP %in% filter_FTO$refsnp_id] <- "FTO"

# color labeling
my_data_harm$fto_col_annot <- "No FTO variant"
my_data_harm$fto_col_annot[my_data_harm$SNP %in% filter_FTO$refsnp_id] <- "FTO variant"

ggplot(data = my_data_harm, 
       mapping = aes(x=explained_variance_BMI*100, y= -log10(p.value.bmi),  
                    col=fto_col_annot, label=gene_label)
       )+
  geom_point(alpha=0.6)+
  ggrepel::geom_text_repel(show.legend = FALSE)+
  xlab("Explained variance in BMI in %")+
  ylab("-log10 P-val")+
  labs(title = "Explained variance exposure",
       col='SNP'
       ) +
  theme_few()+
  scale_colour_few() -> fig00
fig00

# second formula variance explained2
ggplot(data = my_data_harm, 
       mapping = aes(x=explained_variance_BMI2*100, y= -log10(p.value.bmi),  
                    col=fto_col_annot, label=gene_label)
       )+
  geom_point(alpha=0.6)+
  ggrepel::geom_text_repel(show.legend = FALSE)+
  xlab("Explained variance in BMI in %")+
  ylab("-log10 P-val")+
  labs(title = "Explained variance exposure",
       col='SNP'
       ) +
  theme_few()+
  scale_colour_few() -> fig00_b
fig00_b

filenamefig00_a <- paste0("../output/plots/", Sys.Date(), "_figure00_a.png")
ggsave(filename = filenamefig00_a, plot = fig00, 
       width = 10, height = 6, 
       units = "in"  # default
       )

filenamefig00_b <- paste0("../output/plots/", Sys.Date(), "_figure00_b.png")
ggsave(filename = filenamefig00_b, plot = fig00_b, 
       width = 10, height = 6, 
       units = "in"  # default
       )
```

@@

```{r get-data-for-fig1, warning=FALSE}
# 4.       Exclude the IV if Pval > genome-wide significance threshold  (5 × 10−8) **NEW 5x10^(-6))**
# 5.       Calculate cumulative explained variances
# initialize df
df_cum_expl_var <- tibble(n=n, 
                          cum_expl_varX = rep(NA, length(n)), 
                          cum_expl_varX2 = rep(NA, length(n)), 
                          cum_expl_varU = rep(NA, length(n)), 
                          cum_expl_varY = rep(NA, length(n)),
                          count_IVs = rep(NA, length(n))
)

for (i in 1:length(colnames_pval) ) {
  threshold <- 5*10^(-6)
  select_col <- colnames_pval[i]
  # calculate cum explained variance and count of IVs
  summarydat <- my_data_harm %>% 
    filter(!!sym(select_col) < threshold) %>% 
    select(explained_variance_BMI, explained_variance_BMI2, explained_variance_GSD, explained_variance_GBC) %>% 
    summarise(expl_variances =across(.cols = everything(), function(x) sum(x)),
              count_IVs = n())
  # fill up df
  df_cum_expl_var[i, 2] <- summarydat$expl_variances[1]
  df_cum_expl_var[i, 3] <- summarydat$expl_variances[2]
  df_cum_expl_var[i, 4] <- summarydat$expl_variances[3]
  df_cum_expl_var[i, 5] <- summarydat$expl_variances[4]
  df_cum_expl_var[i, 6] <- summarydat$count_IVs
}
```

```{r plot-fig1, warning=FALSE}
# 6.       Represent figure similar to Fig1 in Chatterjee’s paper

# Dual Y axis: we have 284 IVs, scale them to 5% (y-axis [0;5%]): 3/284 
## https://r-graph-gallery.com/line-chart-dual-Y-axis-ggplot2.html
scale_factor <- 4.5/284
df_cum_expl_var <- df_cum_expl_var %>% 
  mutate(
    cum_expl_varX = cum_expl_varX *100, # in percent 
    cum_expl_varX2 = cum_expl_varX2 *100,
    cum_expl_varU = cum_expl_varU *100,
    cum_expl_varY = cum_expl_varY *100,
    count_IVs_scaled = count_IVs*scale_factor)

# plot_data <- df_cum_expl_var %>% 
#   pivot_longer(cols = c(cum_expl_varX, cum_expl_varU, count_IVs_scaled)) %>% 
#   mutate(name = factor(name, labels = c("Number of IVs", "% var(U)", "% var(X)")))
# 
# ggplot(plot_data, aes(x=log10(n), y=value, col=name))+
#   geom_point(alpha=0.6)+
#   geom_line(alpha=0.6)+
#   xlab("log10(N)")+
#   #ylab("% variance explained by IVs")+
#   labs(title = "Figure 1",
#        col='', 
#        caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
#        ) +
#   scale_y_continuous(
#     # Features of the first axis
#     name = "% variance explained by IVs",
#     # Add a second axis and specify its features
#     sec.axis = sec_axis(~./scale_factor, name="Number of IVs")
#   ) + 
#   theme_few()+
#   scale_colour_few() ->fig01
# fig01
# 
# plot_data_b <- df_cum_expl_var %>% 
#   pivot_longer(cols = c(cum_expl_varX2, cum_expl_varU, count_IVs_scaled)) %>% 
#   mutate(name = factor(name, labels = c("Number of IVs", "% var(U)", "% var(X)")))
# 
# ggplot(plot_data_b, aes(x=log10(n), y=value, col=name))+
#   geom_point(alpha=0.6)+
#   geom_line(alpha=0.6)+
#   xlab("log10(N)")+
#   #ylab("% variance explained by IVs")+
#   labs(title = "Figure 1",
#        col='', 
#        caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
#        ) +
#   scale_y_continuous(
#     # Features of the first axis
#     name = "% variance explained by IVs",
#     # Add a second axis and specify its features
#     sec.axis = sec_axis(~./scale_factor, name="Number of IVs")
#   ) + 
#   theme_few()+
#   scale_colour_few() ->fig01_b
# fig01_b
# 
# ggsave(filename = "../output/plots/figure01.png", plot = fig01_b, 
#        width = 10, height = 6, 
#        units = "in"  # default
#        )
# 
# filenamefig01_a <- paste0("../output/plots/", Sys.Date(), "_figure01_a.png")
# filenamefig01_b <- paste0("../output/plots/", Sys.Date(), "_figure01_b.png")
```

```{r updated-fig01, warning=FALSE}
plot_data1 <- df_cum_expl_var %>% 
  pivot_longer(cols = c(cum_expl_varX, cum_expl_varU, cum_expl_varY)) %>% 
  mutate(name = factor(name, labels = c("% var(U)", "% var(X)", "% var(Y)")),
         name = factor(name, levels = c("% var(X)", "% var(Y)", "% var(U)") )
         )

N_annotation <- c(10000, 100000, 200000, 300000)
## plot
ggplot(plot_data1, aes(x=n, y=value, col=name))+
  geom_point(alpha=0.6)+
  geom_line(alpha=0.6)+
  geom_line(aes(x=n, y=count_IVs_scaled, col="Number of IVs"), linetype="dotdash")+
  scale_x_continuous(name = paste0("Size of sample ", as.roman(1)),  # 2-sample-MR or label "Study Size"
                     labels = comma, 
                     breaks = N_annotation, #c( 10000, n[7:length(n)]),
                     limits = c(10000, 300000))+
  #ylab("% variance explained by IVs")+
  labs(title = "Figure 1",
       col='', 
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  scale_y_continuous(
    # Features of the first axis
    name = "% variance explained by IVs",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./scale_factor, name="Number of IVs")
  ) + 
  theme_few()+
  #scale_colour_few() +
  #scale_color_manual(name = "", values = c("% var(U)" = "red", 
  #                                         "% var(X)" = "blue", 
  #                                         "% var(Y)" = "darkblue", 
  #                                         "Number of IVs" = "black"
  #                                         ))
  #
  scale_color_manual(name = "", values = c("% var(U)" = ggthemes_data$few$colors$Dark[2,2][[1]], 
                                           "% var(X)" = ggthemes_data$few$colors$Dark[3,2][[1]], 
                                           "% var(Y)" = ggthemes_data$few$colors$Dark[4,2][[1]], 
                                           "Number of IVs" = "black"),
                     # order
                     breaks = c( "Number of IVs", "% var(X)", "% var(Y)", "% var(U)")
  ) -> fig01.update

fig01.update
# scale_color_manual(name = "Y series", values = c("Y1" = "darkblue", "Y2" = "red"))

plot_data1b <- df_cum_expl_var %>% 
  pivot_longer(cols = c(cum_expl_varX2, cum_expl_varU, cum_expl_varY)) %>% 
  mutate(name = factor(name, labels = c("% var(U)", "% var(X)", "% var(Y)")),
         name = factor(name, levels = c("% var(X)", "% var(Y)", "% var(U)") )
         )
## plot
ggplot(plot_data1b, aes(x=n, y=value, col=name))+
  geom_point(alpha=0.6)+
  geom_line(alpha=0.6)+
  geom_line(aes(x=n, y=count_IVs_scaled, col="Number of IVs"), linetype="dotdash")+
  scale_x_continuous(name = paste0("Size of sample ", as.roman(1)),  # 2-sample-MR or label "Study Size"
                     labels = comma, 
                     breaks = N_annotation, #c( 10000, n[7:length(n)]),
                     limits = c(10000, 300000))+
  #ylab("% variance explained by IVs")+
  labs(title = "Figure 1",
       col='', 
       caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  scale_y_continuous(
    # Features of the first axis
    name = "% variance explained by IVs",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./scale_factor, name="Number of IVs")
  ) + 
  theme_few()+
  #scale_colour_few() +
  #scale_color_manual(name = "", values = c("% var(U)" = "red", 
  #                                         "% var(X)" = "blue", 
  #                                         "% var(Y)" = "darkblue", 
  #                                         "Number of IVs" = "black"
  #                                         ))
  #
  scale_color_manual(name = "", values = c("% var(U)" = ggthemes_data$few$colors$Dark[2,2][[1]], 
                                           "% var(X)" = ggthemes_data$few$colors$Dark[3,2][[1]], 
                                           "% var(Y)" = ggthemes_data$few$colors$Dark[4,2][[1]], 
                                           "Number of IVs" = "black"),
                     # order
                     breaks = c( "Number of IVs", "% var(X)", "% var(Y)", "% var(U)")
  ) -> fig01b.update
fig01b.update

# SAVE
filenamefig01_a <- paste0("../output/plots/", Sys.Date(), "_figure01_a.png")
filenamefig01_b <- paste0("../output/plots/", Sys.Date(), "_figure01_b.png")
ggsave(filename = filenamefig01_a, plot = fig01.update, 
       width = 10, height = 6, 
       units = "in"  # default
       )

ggsave(filename = filenamefig01_b, plot = fig01b.update, 
       width = 10, height = 6, 
       units = "in"  # default
       )
```

Discussion: *Pleiotropy* with increasing sample size. INSIDE assumption.

@@

```{r MAF-Pval, warning=FALSE}
ggplot(data = my_data_harm, aes(x=eaf.bmi, y=-log10(p.value.bmi)))+
  geom_point(alpha=0.6)+
  geom_hline(yintercept=-log10(median(my_data$p_value_exp)), col="red", alpha=0.6)+
  geom_vline(xintercept=median(my_data$eaf_exp), col="red", alpha=0.6)+
  xlab("Effect allele frequency for exposure")+
  ylab("-log10(P-value) ")+
  labs(title = "MAF vs Pval",
       subtitle = "Define different genetic architectures",
       #col='', 
       #caption = "Relationship between sample size, average number of instrumental variables (IVs) and variance of traits explained by the IVs"
       ) +
  theme_few()+
  scale_colour_few() ->fig02
fig02

# ggsave(filename = "../output/plots/figure02.png", plot = fig02)
```

@@

```{r mafVSbetas}
plot_data_2 <- my_data_harm %>% 
  mutate(maf.bmi = ifelse(eaf.bmi >0.5, 1-eaf.bmi, eaf.bmi)) %>% 
  select(maf.bmi, beta.bmi)
